Eco-evolutionary dynamics of finite populations from first principles

### 2023-04-23 ind_files/03_BD_processes.tex
Minor additions and rephrasinf

In appendix \ref{App_density_to_freq}, I use a multivariate version of It\^{o}'s formula to derive a general stochastic equation for the frequencies of each type in the population. Letting $\overline{w} = \sum w_ip_i$ and $\overline{\tau} = \sum \tau_i p_i$ be the average population fitness and turnover respectively, I show in appendix \ref{App_density_to_freq} that the frequency of the $i\textsuperscript{th}$ type in the population $\mathbf{x}(t)$ changes according to the equation:
\begin{equation}
\label{nD_eqn_for_frequencies}
\begin{aligned}
dp_i(t) &= \left[(w_i(\mathbf{x}) - \overline{w})p_i + \mu\left\{Q_i(\mathbf{p}) - p_i\left(\sum\limits_{j=1}^{m}Q_j(\mathbf{p})\right)\right\}\right]dt\\
&- \frac{1}{K}\frac{1}{N_{K}(t)}\left[(\tau_i(\mathbf{x}) - \overline{\tau})p_i + \mu\left\{Q_i(\mathbf{p}) - p_i\left(\sum\limits_{j=1}^{m}Q_j(\mathbf{p})\right)\right\}\right]dt\\
&+ \frac{1}{\sqrt{K}}\frac{1}{N_{K}(t)}\left[\left(A^{+}_{i}(\mathbf{x})\right)^{1/2}dW^{(i)}_t - p_i\sum\limits_{j=1}^{m}\left(A^{+}_{j}(\mathbf{x})\right)^{1/2}dW^{(j)}_t\right]
\end{aligned}
\end{equation}
where $W^{(1)}_t,W^{(2)}_t, \ldots, W^{(m)}_t$ are $m$ independent one-dimensional Wiener processes and I have used the notation $Q_i(\mathbf{p}) = Q_i(\mathbf{x})/N_K(t)$ for notational clarity. I will show below that the first term in this expression describes directional changes in the population composition due to `classical' evolutionary forces such as selection and mutation that occur in deterministic infinite population models. The second term is an additional directional force on population composition that is only seen in finite populations and can be thought of as a biasing `selection' for reduced turnover rate due to an effect similar to gambler's ruin in probability theory. The consequences of this term, as well as connections with previous studies, are discussed in detail in part \ref{part_summary}. Finally, the last term of equation \eqref{nD_eqn_for_frequencies} describes non-directional stochastic effects due to fluctuations and has a `spreading effect' \textcolor{blue}{(Also see equation} \eqref{nD_stoch_RM_noise_term_alt_representation} \textcolor{blue}{in Appendix} \ref{App_alternative_representation} \textcolor{blue}{for a more elegant representation of these noise terms as an integral against a single Wiener process)}. \textcolor{blue}{For the case $m=2, \mu = 0$ (two interacting types of individuals, no mutation), this expression, after using the representation of noise terms presented in Appendix} \ref{App_alternative_representation} \textcolor{blue}{and letting $p = p_1$, becomes the considerably simpler expression
\begin{equation}
	\label{two_species_stoch_RM}
	\begin{aligned}
		dp &= \left[(w_1 - w_2)p(1-p) - \frac{1}{KN_K}(\tau_1-\tau_2)p(1-p)\right]dt\\
		&+ \frac{1}{\sqrt{KN_K}}\left(p(1-p)\left[\tau_1 + (\tau_2 - \tau_1)p\right]\right)^{1/2}dW_t
	\end{aligned}
\end{equation}
It is easy to see that if we start with an initial condition $p_0 \in [0,1]$, the system defined by equation \eqref{two_species_stoch_RM} will always remain in $[0,1]$, and in that sense, it is `well-behaved'. Similar well-behavedness at the boundaries for the more general equation} \eqref{nD_eqn_for_frequencies} \textcolor{blue}{can also be checked directly, and has been carried out in Appendix} \ref{App_well_behaved}.

>>>>

In appendix \ref{App_density_to_freq}, I use a multivariate version of It\^{o}'s formula to derive a general stochastic equation for the frequencies of each type in the population. Letting $\overline{w} = \sum w_ip_i$ and $\overline{\tau} = \sum \tau_i p_i$ be the average population fitness and turnover respectively, I show in appendix \ref{App_density_to_freq} that the frequency of the $i\textsuperscript{th}$ type in the population $\mathbf{x}(t)$ changes according to the equation:
\begin{equation}
\label{nD_eqn_for_frequencies}
\begin{aligned}
dp_i(t) &= \left[(w_i(\mathbf{x}) - \overline{w})p_i + \mu\left\{Q_i(\mathbf{p}) - p_i\left(\sum\limits_{j=1}^{m}Q_j(\mathbf{p})\right)\right\}\right]dt\\
&- \frac{1}{K}\frac{1}{N_{K}(t)}\left[(\tau_i(\mathbf{x}) - \overline{\tau})p_i + \mu\left\{Q_i(\mathbf{p}) - p_i\left(\sum\limits_{j=1}^{m}Q_j(\mathbf{p})\right)\right\}\right]dt\\
&+ \frac{1}{\sqrt{K}}\frac{1}{N_{K}(t)}\left[\left(A^{+}_{i}(\mathbf{x})\right)^{1/2}dW^{(i)}_t - p_i\sum\limits_{j=1}^{m}\left(A^{+}_{j}(\mathbf{x})\right)^{1/2}dW^{(j)}_t\right]
\end{aligned}
\end{equation}
where $W^{(1)}_t,W^{(2)}_t, \ldots, W^{(m)}_t$ are $m$ independent one-dimensional Wiener processes and I have used the notation $Q_i(\mathbf{p}) = Q_i(\mathbf{x})/N_K(t)$ for notational clarity. I will show below that the first term in this expression describes directional changes in the population composition due to `classical' evolutionary forces such as selection and mutation that occur in deterministic infinite population models. The second term is an additional directional force on population composition that is only seen in finite populations and can be thought of as a biasing `selection' for reduced turnover rate due to an effect similar to gambler's ruin in probability theory. The consequences of this term, as well as connections with previous studies, are discussed in detail in part \ref{part_summary}. Finally, the last term of equation \eqref{nD_eqn_for_frequencies} describes non-directional stochastic effects due to fluctuations and has a `spreading effect' \textcolor{blue}{(Also see equation} \eqref{nD_stoch_RM_noise_term_alt_representation} \textcolor{blue}{in Appendix} \ref{App_alternative_representation} \textcolor{blue}{for a more elegant representation of these noise terms as an integral against a single Wiener process)}. \textcolor{blue}{ For the case $m=2, \mu = 0$ (two interacting types of individuals with no mutations between types), after using the representation of noise terms presented in Appendix \ref{App_alternative_representation} and letting $p = p_1$, equation} \eqref{nD_eqn_for_frequencies} \textcolor{blue}{becomes the considerably simpler expression
\begin{equation}
	\label{two_species_stoch_RM}
	\begin{aligned}
		dp &= \left[(w_1 - w_2)p(1-p) - \frac{1}{KN_K}(\tau_1-\tau_2)p(1-p)\right]dt\\
		&+ \frac{1}{\sqrt{KN_K}}\left(p(1-p)\left[\tau_1 + (\tau_2 - \tau_1)p\right]\right)^{1/2}dW_t
	\end{aligned}
\end{equation}
It is easy to see that if we start with an initial condition $p_0 \in [0,1]$, the system defined by equation \eqref{two_species_stoch_RM} will always remain in $[0,1]$, and in that sense, it is `well-behaved'. Similar well-behavedness at the boundaries for the more general equation} \eqref{nD_eqn_for_frequencies} \textcolor{blue}{can also be checked directly, and has been carried out in Appendix} \ref{App_well_behaved}.

### 2023-04-23 ind_files/06_nontechnical_discussion.tex
Minor additions and rephrasinf

In this thesis, I present a mathematical framework for such a reformulation. Part \ref{part_theory} of this thesis develops a formalism for both populations of individuals that vary in discrete characters (Chapter \ref{chap_BD}) as well as populations of individuals that vary in a single one-dimensional quantitative character (Chapter \ref{chap_infD_processes}). The central result of this reformulation is a series of stochastic differential equations derived in Chapter \ref{chap_unification}. To begin, I derive an equation for change in type frequencies in the population (equation \eqref{nD_stochastic_RM}) that generalizes the replicator-mutator equation to finite, fluctuating, closed populations evolving in continuous time. From this, I show how one can derive an equation for changes in the population mean of an arbitrary type-level quantity (equation \eqref{nD_stochastic_Price}) that generalizes the dynamic, continuous time version of the Price equation to finite, fluctuating populations. I also derive an equation for changes in the population variance of an arbitrary type-level quantity (equation \eqref{nD_stochastic_Price_variance}) in such populations that generalizes a recent infinite population formulation of the same~\citep{lion_theoretical_2018}. My work thus generalizes some fundamental formal structures of eco-evolutionary population dynamics to finite, fluctuating populations. In particular, my equations generalize the unifying formalism described in~\cite{lion_theoretical_2018}\footnote{which itself is a reformulation of the relatively well-known unification of eco-evolutionary dynamics via the Price equation~\citep{frank_natural_2012, queller_fundamental_2017, luque_mirror_2021} in a dynamically sufficient, continuous time framework} to finite, fluctuating populations - taking $K \to \infty$ in equations \eqref{nD_stochastic_RM}, \eqref{nD_stochastic_Price}, and \eqref{nD_stochastic_Price_variance} recover equations (6), (11), and (14) in~\cite{lion_theoretical_2018} respectively as their infinite population limits. While my equations thus recover standard results such as the Price equation and the replicator-mutator equation in the infinite population limit, they also predict that these results do not completely capture the behavior of finite populations.

@@

\textcolor{blue}{Note that types that tend to increase the \emph{total} population size $KN_K(t)$ (such as altruists in evolutionary theory and mutualists in ecological communities) will reduce the magnitude of this effect compared to types that do not facilitate such an increase, such as cheaters and highly competitive species. Further, if altruists/mutualists act by reducing the death rate (rather than increasing the birth rate) of other individuals, their presence causes higher $w$ and lower $\tau$ in the beneficiary individuals, both of which are favored by selection (but note that if they act by increasing the birth rate, they increase the magnitude of negative noise-induced selection disfavouring the beneficiary individual), which explains why this effect preferentially favors mutualists in reversing the direction of deterministic selection in finite population models with fluctuating population sizes if interaction effects are on the death rate~\citep{mcleod_social_2019}. Thus, selection for reduced turnover rate could help explain why cooperation often persists in fluctuating populations in laboratory experiments and finite population IbMs of social evolution that allow for fluctuating populations despite infinite population models predicting their extinction~\citep{houchmandzadeh_selection_2012,sanchez_feedback_2013,houchmandzadeh_fluctuation_2015,chotibut_evolutionary_2015,behar_fluctuations-induced_2016,mcavoy_public_2018,mcleod_social_2019}}.

@@

It is worth noting that the existence of noise-induced selection directly implies (from equation \eqref{nD_stochastic_RM} or \eqref{nD_stochastic_Price}) that evolution is not expected to maximize fitness in finite populations even if fitness is entirely frequency independent as long as there is some (heritable) variation in the turnover rates $\tau_i$, further underscoring the now well-appreciated point that the view of evolution as `climbing a hill' on a fitness landscape and thereby maximizing fitness is rather limited~\citep{grodwohl_theory_2017}. The equations of section \ref{sec_fun_theorems} also imply that for the evolution of a trait to be \emph{truly} neutral in finite populations (in the sense of all $m$ types in a system having equal fixation/extinction probability if we start with an initial state in which every type has frequency $1/m$), it is not sufficient for the trait in question to be neutral with respect to fitness $w$. Instead, we also require the trait to be neutral with respect to turnover rate $\tau$. Though this is clear from examining equations \eqref{nD_stochastic_RM} or \eqref{nD_stochastic_Price}, I also illustrate a deviation from neutrality explicitly for the $m=2$ case using a simple resource-competition model in Appendix \ref{App_deviation_from_neutrality}. Indeed, in ecological models, previous work in the `quasi-neutral' regime has shown that in finite, fluctuating populations, the equal growth rate of types is not sufficient to ensure equal fixation probabilities and that there is a slight biasing for types with lower turnover rates, sometimes interpreted as a selection `for longevity'~\citep{lin_features_2012, oliveira_advantage_2017,balasekaran_quasi-neutral_2022}. In models of evolutionary game theory in fluctuating finite populations, individuals with lower death rates have higher fixation probability even when growth rates are equalized~\citep{huang_stochastic_2015, czuppon_fixation_2018}. Similarly, models of cell cycle dynamics find that selection favors cell types that periodically arrest their cell cycle relative to non-arresting cells even when their growth rates are equal~\citep{wodarz_effect_2017}. In the language of the birth-death formalism I develop in this thesis, all of these studies equalize the growth rates $w$ of competing types but allow the turnover to vary (by arresting the cell cycle and thus reducing turnover, for example), thus allowing noise-induced selection for reduced turnover to operate in the system.

Since noise-induced selection is directional, the seemingly small systematic deviations from neutrality exhibited by these models can have significant consequences over long evolutionary times (For example, see~\cite{veller_drift-induced_2017}). My derivations show analytically that such deviations from neutrality in finite populations are a generic phenomenon explained by noise-induced selection and should be expected whenever there is variation in turnover rates. Selection on turnover rates also leads to insights on life-history evolution, and these insights have been extensively reported in a recent pre-print that independently arrives at my equations for type frequencies (equation \eqref{nD_stochastic_RM}) and the change of mean fitness and turnover in the population (equations \eqref{nD_stochastic_Fisher} and \eqref{nD_stochastic_Fisher_turnover} respectively) using certain discrete time stochastic processes and their approximation via techniques reminiscent of numerical stochastic integration~\citep{kuosmanen_turnover_2022}.

@@

@@

A recent study showed (among other things) that in a very broad class of density-dependent competition models, model-specific details can be ignored in favor of a universal `coarse-grained' description that only includes a small number of quantitities~\citep{mazzolini_universality_2022}. Among the generic predictions of this general description was the presence of noise-induced selection in general competition models. Our work generalizes this particular result of~\cite{mazzolini_universality_2022} to models with arbitrary interaction types and presents the relevant equations in a formalism that is more in accordance with standard biological models such as the Price equation. Our results also show that equations like the replicator-mutator equation and Price equation are asymptotically `universal', in the sense that an arbitrary density-dependent birth-death process with functional forms given by \eqref{nD_functional_forms_for_replicator} will satisfy the replicator-mutator or Price equation as we move to infinitely large populations ($K \to \infty$). Thus, these equations provide coarse-grained `fundamental theorems' that are always satisfied for biological populations that are sufficiently large.

~\citet{lion_theoretical_2018} has pointed out that in the dynamic setting (for infinite populations), the replicator-mutator equation \eqref{nD_replicator_mutator} is in some sense the `\textcolor{red}{\st{most}} fundamental' equation, and equations like the dynamically sufficient version of the Price equation are best viewed as a hierarchy of moment equations for the population mean, population variance, etc. of a type-level quantity. This is also true in my framework - equation \eqref{nD_stochastic_RM} is the most fundamental equation for population dynamics, and equations like \eqref{nD_stochastic_Price} and \eqref{nD_stochastic_Price_variance} can then be derived from \eqref{nD_stochastic_RM} through repeated application of It\^o's formula, in principle for any moment of the distribution of the quantity $f$ in the population (though this quickly becomes too tedious to actually carry out in practice). If we additionally assume that the quantity $f$ follows a Gaussian distribution in the population, then the mean and variance completely characterize the distribution of $f$, and equations \eqref{nD_stochastic_RM}, \eqref{nD_stochastic_Price}, and \eqref{nD_stochastic_Price_variance} together specify the complete stochastic dynamics of the system. Note that even in this case, actually solving these equations analytically for equilibrium/stationary state distributions of $\mathbf{p}$, $\overline{f}$, and $\sigma^2_f$ will quickly become impossibly difficult if the birth and death rate functions are complicated. Indeed, previous studies indicate that in high dimensions, evolutionary birth-death models can exhibit a dizzying array of complicated behavior, including limit cycles and chaotic trajectories~\citep{doebeli_diversity_2017}. However, while this may present impediments to formulating exact solutions to the equations I formulate, the point of these equations is not necessarily to solve them to begin with.

Indeed, the most important takeaway from this thesis is not in the solutions to the equations of section \ref{sec_fun_theorems} but in the formulation and the equations themselves. The equations I derive are very general, since part \ref{part_theory} makes essentially no assumptions other than density dependence, the impossibility of infinite growth starting from finite population size, and the ability to define per-capita birth and death rates, and thus, these birth and death rate function(al)s can in general be almost arbitrarily complicated. They thus apply to a very wide array of biological populations. Further, as we saw in section \ref{sec_fun_theorems}, the terms of these equations lend themselves to simple biological interpretation and make some general \emph{qualitative} predictions about how evolution should operate in finite, fluctuating populations. Namely, my equations show that evolution should operate similarly to how it operates in infinite populations, with the addition of some non-directional fluctuations (this is just drift) as well as an extra \emph{directional} force (noise-induced selection). Like the classical Price equation, the utility of the equations of section \ref{sec_fun_theorems} thus lies not (necessarily) in their solutions for specific models, but instead in their generality and the fact that their terms help us clearly think about the various evolutionary phenomena at play in biological populations~\citep{frank_natural_2012,luque_one_2017, luque_mirror_2021}. The general spirit of this thesis is thus in line with the general idea of trying to formulate `model-independent' eco-evolutionary theory that has recently been (rapidly) gaining popularity in the literature~\citep{grafen_formal_2014, queller_fundamental_2017, lion_theoretical_2018, allen_mathematical_2019, rice_universal_2020, week_white_2021, wickman_theoretical_2022, kuosmanen_turnover_2022, mazzolini_universality_2022,lion_extending_2023,allen_natural_2023}.

>>>>

In this thesis, I present a mathematical framework for such a reformulation. Part \ref{part_theory} of this thesis develops a formalism for both populations of individuals that vary in discrete characters (Chapter \ref{chap_BD}) as well as populations of individuals that vary in a single one-dimensional quantitative character (Chapter \ref{chap_infD_processes}). The central result of this reformulation is a series of stochastic differential equations derived in Chapter \ref{chap_unification}. To begin, I derive an equation for change in type frequencies in the population (equation \eqref{nD_stochastic_RM}) that generalizes the replicator-mutator equation to finite, fluctuating, closed populations evolving in continuous time. From this, I show how one can derive an equation for changes in the population mean of an arbitrary type-level quantity (equation \eqref{nD_stochastic_Price}) that generalizes the dynamic, continuous time version of the Price equation to finite, fluctuating populations. I also derive an equation for changes in the population variance of an arbitrary type-level quantity (equation \eqref{nD_stochastic_Price_variance}) in such populations that generalizes a recent infinite population formulation of the same~\citep{lion_theoretical_2018}. My work thus generalizes some fundamental formal structures of eco-evolutionary population dynamics to finite, fluctuating populations. \textcolor{red}{\st{In particular, my equations generalize the unifying formalism described in Lion, 2018 to finite, fluctuating populations - taking $K \to \infty$ in equations (5.1), (5.2), and (5.9) recover equations (6), (11), and (14) in Lion, 2018 respectively as their infinite population limits}} \textcolor{blue}{The equations I derive deal with biologically important quantities in a `coarse-grained' manner, lend themselves to simple biological interpretation, and are very general. They thus fulfill the criteria to be called `fundamental theorems' (in the sense of~\cite{queller_fundamental_2017}) and present a `unifying perspective' (in the sense of~\cite{lion_theoretical_2018}) for the dynamics of finite populations.} While my equations thus recover standard results such as the Price equation and the replicator-mutator equation in the infinite population limit, they also predict that these results do not completely capture the behavior of finite populations.

@@

\textcolor{blue}{Populations consisting mostly of types that tend to increase the \emph{total} population size $KN_K(t)$ (such as altruists in evolutionary theory and mutualists in ecological communities) will experience a reduced magnitude of noise-induced selection acting in the system compared to Populations consisting mostly of types that do not facilitate such an increase, such as cheaters and highly competitive species. Further, if altruists/mutualists act by reducing the death rate (rather than increasing the birth rate) of other individuals, their presence causes higher $w$ and lower $\tau$ in the beneficiary individuals, both of which are favored by evolution through classical selection and noise-induced selection respectively (but note that if they act by increasing the birth rate, they increase the magnitude of negative noise-induced selection disfavouring the beneficiary individual). This is why the direction of deterministic selection is specifically reversed to favor mutualists in finite population models with fluctuating population sizes if interaction effects are on the death rate~\citep{mcleod_social_2019}. Thus, selection for reduced turnover rate could help explain why cooperation often persists in fluctuating populations in laboratory experiments~\citep{sanchez_feedback_2013} and finite population IbMs~\citep{houchmandzadeh_selection_2012,,houchmandzadeh_fluctuation_2015,chotibut_evolutionary_2015,behar_fluctuations-induced_2016,mcavoy_public_2018,mcleod_social_2019} of social evolution despite infinite population models predicting their extinction}.

@@

It is worth noting that the existence of noise-induced selection directly implies (from equation \eqref{nD_stochastic_RM} or \eqref{nD_stochastic_Price}) that evolution is not expected to maximize fitness in finite populations even if fitness is entirely frequency independent as long as there is some (heritable) variation in the turnover rates $\tau_i$, further underscoring the now well-appreciated point that the view of evolution as `climbing a hill' on a fitness landscape and thereby maximizing fitness is rather limited~\citep{grodwohl_theory_2017}. The equations of section \ref{sec_fun_theorems} also imply that for the evolution of a trait to be \emph{truly} neutral in finite populations (in the sense of all $m$ types in a system having equal fixation/extinction probability if we start with an initial state in which every type has frequency $1/m$), it is not sufficient for the trait in question to be neutral with respect to fitness $w$. Instead, we also require the trait to be neutral with respect to turnover rate $\tau$. Though this is clear from examining equations \eqref{nD_stochastic_RM} or \eqref{nD_stochastic_Price}, I also illustrate a deviation from neutrality explicitly for the $m=2$ case using a simple resource-competition model in Appendix \ref{App_deviation_from_neutrality}. Indeed, in ecological models, previous work in the `quasi-neutral' regime has shown that in finite, fluctuating populations, the equal growth rate of types is not sufficient to ensure equal fixation probabilities and that there is a slight biasing for types with lower turnover rates, sometimes interpreted as a selection `for longevity'~\citep{lin_features_2012, oliveira_advantage_2017,balasekaran_quasi-neutral_2022}. In models of evolutionary game theory in fluctuating finite populations, individuals with lower death rates have higher fixation probability even when growth rates are equalized~\citep{huang_stochastic_2015, czuppon_fixation_2018}. Similarly, models of cell cycle dynamics find that selection favors cell types that periodically arrest their cell cycle relative to non-arresting cells even when their growth rates are equal~\citep{wodarz_effect_2017}. In the language of the birth-death formalism I develop in this thesis, all of these studies equalize the growth rates $w$ of competing types but allow the turnover to vary (by arresting the cell cycle and thus reducing turnover, for example), thus allowing noise-induced selection for reduced turnover to operate in the system. Since noise-induced selection is directional, the seemingly small systematic deviations from neutrality exhibited by these models can have significant consequences over long evolutionary times (For example, see~\cite{veller_drift-induced_2017}). My derivations show analytically that such deviations from neutrality in finite populations are a generic phenomenon explained by noise-induced selection and should be expected whenever there is variation in turnover rates. Selection on turnover rates also leads to insights on life-history evolution, and these insights have been extensively reported in a recent pre-print~\citep{kuosmanen_turnover_2022} \textcolor{red}{\st{that independently arrives at my equations for type frequencies (equation (5.1)) and the change of mean fitness and turnover in the population (equations (5.7) and (5.8) respectively) using certain discrete time stochastic processes and their approximation via techniques reminiscent of numerical stochastic integration (Kuosmanen et al., 2022).}}.

@@

\section{Connections with other theoretical frameworks}

\textcolor{blue}{~\cite{lion_theoretical_2018} has recently proposed a reformulation of the relatively well-known unification of eco-evolutionary dynamics via the Price equation~\citep{frank_natural_2012, queller_fundamental_2017, luque_mirror_2021} in a dynamically sufficient, continuous time framework using `feedback' as a `unifying perspective'. Our equations generalize the unifying framework described in~\cite{lion_theoretical_2018} to finite, fluctuating populations --- taking $K \to \infty$ in equations \eqref{nD_stochastic_RM}, \eqref{nD_stochastic_Price}, and \eqref{nD_stochastic_Price_variance} recover equations (6), (11), and (14) in~\cite{lion_theoretical_2018} respectively. \cite{lion_theoretical_2018} has pointed out that in the dynamic setting (for infinite populations), the replicator-mutator equation \eqref{nD_replicator_mutator} is in some sense the `most fundamental' of the lot, and equations like the Price equation are best viewed as a hierarchy of moment equations for the population mean, population variance, etc. of a type-level quantity. This is also true in our framework - equation \eqref{nD_stochastic_RM} is the fundamental equation for population dynamics, and equations like \eqref{nD_stochastic_Price} and \eqref{nD_stochastic_Price_variance} can then be derived from \eqref{nD_stochastic_RM} through repeated application of It\^o's formula (in principle for any moment of the distribution of $f$ in the population, though this quickly becomes too tedious to actually carry out in practice). If we assume that the quantity $f$ follows a Gaussian distribution, then the mean and variance completely characterize the distribution,  and equations \eqref{nD_stochastic_RM}, \eqref{nD_stochastic_Price}, and \eqref{nD_stochastic_Price_variance} together specify the complete stochastic dynamics of the system.}

\textcolor{blue}{The equivalent of our stochastic equations has recently been derived for quantitative traits from a very different starting point using the theory of measure-valued branching processes~\citep{week_white_2021} --- Equations (21b) and (21c) in \cite{week_white_2021} are exactly the $m \to \infty$ version of our equations for changes in the mean value of a type-level quantity and changes in the variance of a type-level quantity respectively for the special case in which the type-level quantity is the value of the quantitative trait being studied. The methods used by these authors is exactly complementary to the field equations we formulate in Chapter~\ref{chap_infD_processes} (See section~\ref{sec_disc_field_eqns} for a more detailed technical discussion). A recent preprint has also independently arrived at our equations for type frequencies (equation \eqref{nD_stochastic_RM}) and the change of mean fitness and turnover in the population (equations \eqref{nD_stochastic_Fisher} and \eqref{nD_stochastic_Fisher_turnover}) in the context of life-history evolution using certain discrete time stochastic processes and their approximation via techniques reminiscent of numerical stochastic integration~\citep{kuosmanen_turnover_2022}.}

\textcolor{blue}{A recent study showed that in a very broad class of density-dependent competition models, model-specific details can be ignored in favor of a universal `coarse-grained' description that only includes a small number of very generally defined quantities~\citep{mazzolini_universality_2022}. Among the generic predictions of this general description was the presence of noise-induced selection in general competition models. Our work generalizes this particular result of~\cite{mazzolini_universality_2022} to models with arbitrary interaction types and presents the relevant equations in a formalism that is more in accordance with standard biological models such as the Price equation. Our results also show that equations like the replicator-mutator equation and Price equation are asymptotically `universal', in the sense that an arbitrary density-dependent birth-death process with functional forms given by \eqref{nD_functional_forms_for_replicator} will satisfy the replicator-mutator or Price equation as we move to infinitely large populations ($K \to \infty$). Thus, these equations provide coarse-grained `fundamental theorems' that are always satisfied for biological populations that are sufficiently large.}

@@

Actually solving these equations analytically for equilibrium/stationary state distributions of $\mathbf{p}$, $\overline{f}$, and $\sigma^2_f$ will likely quickly become impossibly difficult if the birth and death rate functions are complicated. Indeed, previous studies indicate that in high dimensions, evolutionary birth-death models can exhibit a dizzying array of complicated behavior, including limit cycles and chaotic trajectories~\citep{doebeli_diversity_2017}. However, while this may present impediments to formulating exact solutions to the equations I formulate, the point of these equations is not necessarily to solve them to begin with. Indeed, the most important takeaway from this thesis is not in the solutions to the equations of section \ref{sec_fun_theorems} but in the formulation and the equations themselves. The equations I derive are very general, since part \ref{part_theory} makes essentially no assumptions other than density dependence, the impossibility of infinite growth starting from finite population size, and the ability to define per-capita birth and death rates, and thus, these birth and death rate function(al)s can in general be almost arbitrarily complicated. They thus apply to a very wide array of biological populations. Further, as we saw in section \ref{sec_fun_theorems}, the terms of these equations lend themselves to simple biological interpretation and make some general \emph{qualitative} predictions about how evolution should operate in finite, fluctuating populations. Namely, my equations show that evolution should operate similarly to how it operates in infinite populations, with the addition of some non-directional fluctuations (this is just drift) as well as an extra \emph{directional} force (noise-induced selection). Like the classical Price equation, the utility of the equations of section \ref{sec_fun_theorems} thus lies not (necessarily) in their solutions for specific models, but instead in their generality and the fact that their terms help us clearly think about the various evolutionary phenomena at play in biological populations~\citep{frank_natural_2012,luque_one_2017, luque_mirror_2021}. The general spirit of this thesis is thus in line with the general idea of trying to formulate `model-independent' eco-evolutionary theory that has recently been (rapidly) gaining popularity in the literature~\citep{grafen_formal_2014, queller_fundamental_2017, lion_theoretical_2018, allen_mathematical_2019, rice_universal_2020, week_white_2021, wickman_theoretical_2022, kuosmanen_turnover_2022, mazzolini_universality_2022,lion_extending_2023,allen_natural_2023}.
